---
title: "Exercise 4: Numbers of police"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Instructions:

This document contains information, questions, R code, and plots.

**Hints and reminders are bold**

<span style="color:blue"> Questions appear in blue. </span>

**Due midnight 25th February 2022**^[if this is after 2022, you will be given extra time to invent a time machine]


### Rationale:

To complete this exercise you will need to use:

* The theory you have been taught in the
lectures
* The R code you have been given in the intro
* The R help pages and google
* Some biological knowledge
* A bit of creativity and problem solving

**This week the exercise looks at how model selection can be important for
fitting the data, and drawing the right conclusions.** 

-----

### R hints and general help

## Resources:

* P60-65 in 'The New Statistics with R'
* Google (or any other search engine) - especially "diagnostic plots in R"
* [Week 6 module](https://www.math.ntnu.no/emner/ST2304/2023v/Module06/ModelCheckingModule.html)

## R this week:

New functions:

* `qqnorm()` this is a function that produces a normal QQ plot 
(quantile-quantile). It
plots the quantiles from your data (here residuals) against those of a 
theoretical perfect normal distribution.
* `qqline()` this function adds a line to the plot from `qqnorm()` of 1 to 
1 match between
your data (residuals) and the theoretical normal distribution. 
* `plot(YOURMODEL, which=4)` this produces a cook's distance plot. `plot(YOURMODEL)` 
by 
default produces 4 diagnostic plots. Using the `which=4` argument just gives us the 4th one.
* indexing: to take only parts of an object in R, we use `[ ]`. 
If that object has two dimensions e.g.
rows and columns, we can take part of it like this: `OBJECT[row number, column number]`
* to remove a specific row or column put a `-` in front of the number or name. E.g. `OBJECT[-1,]`
removes the first row of the object. 
* `MASS::boxcox()` this code uses the `boxcox()` function within the MASS package. It can
be used to plot the log-likelihood for parameters based on different power transformations. 
* extracting R-squared: `summary(model)$r.squared`

Things to remember:

* `lm()` takes the arguments x, y, and data e.g. `lm(y ~ x, data = YOURDATA)` 
x and y here correspond to an 
explanatory variable (x) and a response (y)
* `plot()` and `abline()` **hint: you can use `abline(a=intercept, b=slope)` OR `abline(YOURMODEL)`**
* `install.packages()` **if you can't remember go [here](https://www.math.ntnu.no/emner/ST2304/2023v/RIntro/R_tutorial.html)** Part F.
* `predict()` **hint: remember to use `lm()` with a data argument**

-----

## The challenge: How many police officers do you need?

You are part of the police headquarters team in Chicago. 
It is predicted to be a very cold weekend (not that cold compared to the 
last few weeks in Trondheim! But still cold globally), with an average temperature of
-10&deg;C. Obviously most of the police officers would like to make the most
of the cold weather by going skiing and ice skating. But you also need to 
keep the public safe. Previous research has shown that approximately 2 police
officers (they work in pairs) are needed for every 20 daily crimes. The Chief of Police
has asked your team to provide a recommendation for how many officers
they need for this cold weekend.

Luckily (as always) you already have data on temperature and crime numbers
that you can use.

<span style="color:blue"> **It is your job to find out how many
police officers you would recommend to be on duty on Saturday.** </span>


-----

## Part A: First recommendation

The data from last week's exercise can be found at
 https://www.math.ntnu.no/emner/ST2304/2023v/Module05/NoFreezeTempCrimeChicago.csv

The first step is to import the data and assign it to an object. You can use the whole
web link above to import the data. It is a csv file with column names (header) included.
<details><summary>R hint</summary>
You do not need to download the file first: you can read it into R directly, using `read.csv()`
</details>


The next step is **to plot the data**, its good to remind yourselves what it looked like.

<details><summary>R hint</summary>

```{r PlotData, eval=FALSE}
# Plot the data, include x and y axis labels
plot(YOURDATA$TempC, YOURDATA$Crime, pch=16, 
     xlab="Temp (degree C)",
     ylab="Daily crime number", las=1)

# Run the regression model 
model <- lm(Crime ~ TempC, data=YOURDATA)

# Plot the regression line in dark blue
abline(model, col="darkblue")
```

</details>

<span style="color:blue"> A1. Using the linear regression model
from last week's exercise, **predict** the number of daily crimes for an average 
temperature of -10&deg;C.
Report the answer with prediction interval. </span>

**Hint1: you will need to use the `predict()` function, 
create newdata, and predict with a prediction interval.** If you forget, there
is a reminder below. 

**Hint2: for the lm() you will need to use format lm(y ~ x, data = YourData)**

<details><summary>R hint</summary>

```{r RunModel, eval=FALSE}
# run the model
model1 <- lm(Crime ~ TempC, data=YOURDATA)

# create newdata
newdata <- data.frame(TempC=-10)

# create the predictions
predictions <- predict(model1, newdata, interval="prediction")
```

</details>

<span style="color:blue"> A2. Based on this result
how many police officers would you recommend to be on duty?
**Explain** reasons behind your answer. Include prediction intervals. </span>


-----

## Part B: Model checking

You are speaking to a colleague at lunch, discussing this 
project (because you are very excited about it). Your
colleague tells you they have some extra data. **They 
have mean daily crime numbers from temperatures under 0&deg;C.
This is just what you needed!**

You can find the new complete dataset here:

https://www.math.ntnu.no/emner/ST2304/2023v/Module06/TempCrimeChicago.csv  

It is important to import the data and **plot it**.


<span style="color:blue"> B1. **Fit** a linear regression to the complete
dataset including days <0&deg;C </span> <span style="color:blue"> and
 **plot** the regression line. </span>

<details><summary>R hint</summary>

You should be able to do this using the code in Part A but change the 
dataset and the model name.

</details>

<span style="color:blue"> B2. Look at your plot of the data
and regression line. **What do you think** of the fit?
Are there any problems with using a straight line on this data? 
Do this just by looking.</span>

You have had a go at checking this model just by looking at the data
and the regression line. But there are some more thorough ways you can
explicitly check whether the model meets the assumptions of a linear regression.

The four graphs that statisticians typically use are called:
**Residuals vs fitted, Normal Q-Q, Residuals vs leverage, and cook's distance**

There are some examples of these in the **Model checking** section of the 
linear regression webpage [click here](https://bioceed.uib.no/dropfolder/sites/bioSTATS-and-R/StatisticsInR/2.1_simple_linear_regression.html).

<span style="color:blue"> B3. Create a residuals vs fitted plot for
the linear model on all the data. **Interpret** the plot in terms of
model fit. </span>

Think about which assumption this plot assesses, what would you expect it to look
like if the assumption is met? and how does your plot look different? 

<details><summary>R hint</summary>

```{r ResidPlotHint, eval=FALSE}
# This is some code to help make a residuals vs fitted plot

# Step 1 take out the residuals from the model
# the model is the result of using lm()

# You can also round them
# create a vector of rounded residuals
CrimeResiduals <- round(residuals(model2),2)

# Step 2, take out the fitted values from the model
# create a vector of rounded fit
CrimeFitted <- round(fitted(model2),2)

# plot the fitted and residuals
# Fitted on X, residuals on Y
plot(CrimeFitted, CrimeResiduals)

# add a horizontal line at 0
# line is grey and dashed (lty=2)
abline(h=0, lty=2, col="grey")
```

</details>

<span style="color:blue"> B4. Create a Normal Q-Q plot for
the linear model on all the data. **Interpret** the plot in terms of
model fit. </span>

Again: Think about which assumption this plot assesses, what you expect it to look
like if the assumption is met, and how your data differs from what you expect.

<details><summary>R hint</summary>

```{r QQhint, eval=FALSE}
# use the residuals you calculated above 

# now create the Normal Q-Q plot 
qqnorm(CrimeResiduals)
# add the ideal line
qqline(CrimeResiduals)
```

</details> 

</br>

**You should be able to do this using the R section at beginning of this 
document and the
code for Question B3**. 

<span style="color:blue"> B5. Plot and **Interpret** the Cook's
distance plot for your model. What does it tell you that the Residuals vs fitted and
Normal Q-Q plots did not? </span>

<details><summary>R hint</summary> 



```{r CooksHint, eval=FALSE}
# Code to make cook's distance plot
# this is done by using the plot function (which would
# create 4 plots of the model) and choosing just the
# last one (which=4). A bit of a cheat to a nice plot.

plot(model2, which=4)
```

</details> **But try using the R help at the beginning of this document first!**
    
<span style="color:blue"> B6. Based on your assessments of the model
checking plots, **how** could you improve the model fit for
these data? (suggest at least 2 things you could do - try not to cheat) </span>

If you see your answer to B6 was wrong once you open part C below - don't just
change it, try adding an explanation for why you got it wrong and how seeing
the answer changed what you look for in future. (You could also have noticed
something we missed!)

-----

## Part C: 

<details><summary>Only open once you have answered B6.</summary>

The colleague that gave you the extra data has come back to see how you are 
getting on. They suggest that the main assumption not being met is linearity.
A straight line does not seem to capture the data because it is curved. There 
are also some outliers. The outliers are values of 900 and 1200 
crimes per day for two cold days 
close to -20&deg;C

Your team decides to remove the outliers. You have reason to believe 
they might be typos/incorrect data.

<span style="color:blue"> C1. **What** are some positives and negatives of
removing outliers? e.g. What things should you consider when removing them? </span>

<span style="color:blue"> C2. **Remove** the outliers from your data and plot again. </span>

The outliers are at rows 4 and 7. You can combine these numbers using the function
`c(,)`. If you are unsure what to do, look at the R help section at the
beginning of this document or Google "how to remove rows in R" and see what you get.


The data is still curved. So, you will want to use a transformation
of the response variable or a polynomial (square or cube etc). But which one?

You can use Box-Cox to indicate what
kind of transformation might help with improve the linear regression.
The plot shows the likelihood for different powers of transformation.
E.g. 2 is a squared transformation, 3 is cubic etc.

*R extra information: to ensure you use a function from a particular
package you have loaded you can write the function as `PackageName::FunctionName`.
There is an example of this below.*

```{r BoxCoxHint, eval=FALSE}
# You might need to install the package MASS
# install.packages("MASS")

# Run the boxcox function
# lambda = the power transformation
MASS::boxcox(YOURMODEL, lambda = seq(1,4, length=30))
```

Box-Cox suggests that a quadratic (x<sup>2</sup>)
transformation. You could either transform the response variable
OR add a quadratic term as an explanatory variable. You choose
to try the second and add the quadratic term. 

<span style="color:blue"> C3. **Add** the quadratic term to your model and run again. 
(show code for this answer) </span>

When you add a quadratic variable, remember to keep the original variable in
the model too!

<details><summary>R hint</summary>

To add a quadratic effect you need to use the following format:

`I(variable^2)`, where variable = your variable name e.g. TempC.

You then put this quadratic term after your original variable separated by a +.
e.g. `lm(Y~ X + I(X^2), data = YOURDATA)`

<details><summary>I'm still confused</summary>


```{r QuadHelp, eval=FALSE}
# Create a linear model
# to add a quadratic (or any power) term you must right the
# explanatory variable as I(variable^2) AND keep in the original
# explanatory variable. See example below.
# We need both the linear and quadratic components.

model3 <- lm(Crime ~ TempC + I(TempC^2), data = NoOutliers)
```

</details>
</details>

<span style="color:blue"> C4. **Plot** a new residuals vs fitted plot for 
the new model from C3. </span>






<span style="color:blue"> C5. Look at the new Residuals vs Fitted plot.
**What** do you think of this new model? Has it improved the fit? </span>


</details>


<span style="color:blue"> C6. Now **predict** the number of crimes for a day of -10&deg;C 
from the new model from C3. 
Does this change your recommendation for the number of police needed? If so, **how**? </span>

-----

## Part D: Reflection

<span style="color:blue"> D1. Which model was better at explaining variation in crime? 
Work out the R squared for the models from 
Questions B1 and C3. **How much** of the variance
in daily crime numbers does each model explain? **Compare them.** </span>

The code to do this is at the beginning of this document. 

<span style="color:blue"> D2. Think about the biological context of the results.
**Why** could there be a quadratic relationship between daily crime numbers and
temperature? **How** could you try to find out what the reasons are? E.g.
new studies or data you would need. </span>

-----

## Part E: Feedback

<span style="color:blue"> E1. How do you think this exercise went? What do
you think your group did well, what are you less sure about? 
(2 examples of each) </span>

<span style="color:blue"> E2. What do you think you improved from last 
week? </span>

<span style="color:blue"> E3. Are there any concepts you are very unsure of? </span>

<span style="color:blue"> E4. What would you like feedback on this week? </span>

